Lexical frequency mediates the facilitation effect of cognateness in bilingual lexical acquisition

### 2023-04-17 manuscript.tex
Fix comments and use apaquarto extension

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=20mm,heightrounded]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The role of cross-linguistic lexical similarity on bilingual word acquisition},
  pdfauthor={Gonzalo Garcia-Castro; Daniela S. Ávila-Varela; Ignacio Castillejo; Núria Sebastian-Galles},
  pdfkeywords={lexical acquisition, vocabulary, bilingualism},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{The role of cross-linguistic lexical similarity on bilingual word
acquisition\thanks{This research was supported by grants from the
Spanish Ministerio de Ciencia, Innovación y Universidades
(PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government
{[}ICREA (Catalan Institution for Research and Advanced Studies)
Academia 2019 award{]}. Gonzalo Garcia-Castro was supported by a
fellowship of the Spanish Ministerio de Ciencia, Innovación y
Universidades (FPI 2019). The authors declare no conflicts of interest
with regard to the funding source of this study. We are grateful to
Chiara Santolin, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and
Ege E. Özer for their help-ful feedback. We thank Xavier Mayoral, Silvia
Blanch, and Cristina Cuadrado for their technical support. We also thank
Cristina Dominguez and Katia Pistrin for their efforts in recruiting
infants. We would like to thank the clinics Quirón and Sagrada Familia
that allowed us to recruit participants in their premises. We also thank
all families and infants who participated in the experiments. Data
collection was half-way when the COVID-19 pandemic started. We would
like to pay special tribute to the families that collaborated with us
under these difficult circumstances.}}
\author{Gonzalo Garcia-Castro \and Daniela S. Ávila-Varela \and Ignacio
Castillejo \and Núria Sebastian-Galles}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Bilinguals face the challenging task of learning words from languages
with overlapping phonologies. Floccia et al.~(2018) reported larger
vocabulary sizes for 24-month-old bilinguals that were learning
languages that shared a greater amount of cognates (e.g.,
English-Dutch). The mechanisms underlying this effect remain unknown. We
explore two compatible scenarios. First, we test whether cognates are
learnt earlier than non-cognates. This would account for the difference
in vocabulary size associated to the amount of shared cognates across
languages. Second, we explore the possibility that the word-forms of one
language interact with those form the other language, scaffolding the
acquisition of their translation equivalents when their phonologies
overlap. This mechanism, in line with the parallel activation account of
bilingual speech perception, would provide a plausible explanation to
why cognates are acquired ealier by bilinguals. We developed an online
tool to collect parental reports of receptive and productive
vocabularies from children learning Catalan and/or Spanish, and present
data on receptive and productive vocabulary of bilingual toddlers aged
10 to 36 months.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable]}{\end{tcolorbox}}\fi

\setstretch{1.25}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The foundations of word learning are in place early in age: before the
end of their first year of life infants start directing their gaze to
some objects when hearing their labels, according to both experimental
data (Bergelson \& Swingley, 2012, 2015; Jusczyk \& Aslin, 1995; Tincoff
\& Jusczyk, 1999) and parental reports (e.g., Fenson et al., 2007).
During the last half of their second year, they acquire new words at an
increasingly fast rate (Bergelson, 2020; Bloom, 2002; Fenson et al.,
1994; Goldfield \& Reznick, 1990; Mayor \& Plunkett, 2011). These early
stages of lexical acquisition are characterised by substantial variation
across children, reflected, for instance, on the variation on the number
of words they know (i.e.~vocabulary size, Fenson et al., 1994; Frank et
al., 2021) or on the proportion of those words that fall into the
category of nouns, as opposed to verbs, adjectives, or function words
(e.g., Nelson, 1973, p. bates1994developmental).

Despite this variability, children's trajectories of vocabulary growth
seem quite stable across languages. Tardif et al. (2008) collected data
about the first ten words acquired by 10 to 16 month-old infants living
in the United States, Hong Kong, and Beijing. Since birth, these infants
had been learning English, Cantonese and Mandarin, respectively. The
authors found a common pattern across the three groups: their first ten
words referred to roughly the same concepts, namely relatives/caretakers
(\emph{daddy}, \emph{mommy}), social routines (\emph{bye}, \emph{uh-oh})
or animals (\emph{woof-woof}). These results were later extended by
Frank et al. (2021) to a wider diversity languages, also reporting that
such cross-language commonalities are stronger at earlier stages of
lexical acquisition, compared to later stages. Most of the literature on
early word acquisition, however, has been conducted on monolingual
children, and neglects the problem of how bilinguals--who represent a
substantial proportion of the population in most societies--acquire
words at early ages.

There is evidence that bilinguals know, on average, less words in each
of their languages than monolinguals, but also that both groups know a
similar amount of words when the two languages are aggregated. For
example, Hoff et al. (2012) found that English-Spanish bilingual
toddlers in South Florida knew less words in English than monolinguals,
who only learnt English. Both groups knew a similar total amount of
words when both English and Spanish vocabularies were counted together,
pointing to the importance of collecting data on both languages when
assessing bilinguals' communicative development. Other studies have
provided converging evidence that bilinguals know a similar--or even
larger--number of words than monolinguals, only when the languages are
aggregated (Gonzalez-Barrero et al., 2020; Oller \& Eilers, 2002; J.
Patterson et al., 2004; J. L. Patterson, 2004; Pearson et al., 1993;
Pearson \& Fernández, 1994; Petitto et al., 2001; Smithson et al.,
2014). While these studies have mostly relied on samples of bilingual
children learning two relatively distant languages, as it is the case of
English and Spanish, it is unclear whether children learning
typologically more similar languages also know less words in each of
their languages than monolinguals. What role could linguistic distance
play during early vocabulary growth?

For a given set of concepts, bilingual children may be exposed to two
distinct sets of word-forms--one in each language. Depending on the
linguistic distance between both languages, the two sets of words may
overlap in varying degrees. Particularly, when both languages are
typologically close, like Spanish and Catalan (both Roman languages),
they are more likely to share a large amount of cognates (i.e.,
form-similar translation equivalents) than two linguistically distant
languages, like Spanish and English (one Roman, the other Germanic). For
instance, in the presence of a door, a Spanish-Italian (or a
Spanish-Catalan) bilingual might hear \emph{puerta} and \emph{porta}
(cognates), whereas a Spanish-English bilingual might hear \emph{puerta}
and \emph{door} (non-cognates). It could be the case that mapping two
phonologically similar labels (cognates like \emph{puerta}-\emph{porta})
onto the same referent is easier than doing the same with two
phonologically dissimilar labels (non-cognates, like \emph{puerta} and
\emph{door}). If cognates are easier to acquire than non-cognates,
bilinguals learning a pair of languages that share a high proportion of
cognates should benefit more often from this facilitation effect than
those learning a pair of languages with a lower proportion of cognates,
and should therefore show larger vocabulary sizes.

Floccia et al. (2018) provided evidence in line with this claim. The
authors collected vocabulary data on word comprehension and production
from 372 24-month-old bilingual toddlers living in the United Kingdom
who were learning English and an additional language. The additional
language was one a pool of 13 typologically diverse languages: Bengali,
Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian,
Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors
calculated the average phonological similarity between the words in each
of these additional languages and their translation equivalents in
English. Phonological similarity was measured by computing the
Levenshtein distance between each cross-language pair of phonological
transcriptions. The Levenshtein distance is a metric that computes the
edit distance between two strings by counting the smallest number of
insertions, deletions and substitutions one of the strings has to go
through to become identical to the other (Levenshtein et al., 1966). The
resulting scores were then divided by the length of the longest string
to bound the similarity scores between 0 and 1, and then entered this
variable as a predictor as they modelled participants' vocabulary sizes.
Among other findings, the authors reported an increase in productive
vocabulary size in the additional language associated with an increase
in the average phonological similarity between the translation
equivalents of each language pair. For example, English-Dutch bilinguals
(22.14\% phonological similarity), were able to produce more Dutch words
than English-Mandarin bilinguals (1.97\% phonological similarity) were
able to produce in Mandarin.

Floccia et al.~pointed to \emph{parallel activation} as the main
mechanism underpinning their results. The parallel activation hypothesis
suggests that bilinguals activate both languages simultaneously during
speech production or comprehension, and that this phenomenon is the
result of the activation of lexical representations in both languages,
even when only one is in use during production (Costa et al., 2000;
Hoshino \& Kroll, 2008) or comprehension (Spivey \& Marian, 1999;
Thierry \& Wu, 2007). One of the clearest examples of parallel
activation was provided by Costa et al. (2000). In this study,
Catalan-Spanish monolingual and bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trials
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}). Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
Importantly, Spanish monolinguals--unfamiliar with the Catalan
translations of the Spanish words they uttered--did not show this
effect. These results suggest that bilinguals' Catalan phonology was
activated during the production of Spanish words, facilitating the
naming of cognate pictures. Several subsequent studies have also
provided similar evidence in comprehension in children (Poulin-Dubois et
al., 2013; e.g., Von Holzen \& Mani, 2012). Parallel activation is
therefore a plausible mechanism to account for Floccia et al.'s results:
cognates increase the amount of activation in both languages,
facilitating word acquisition, and ultimately leading to children
learning language pairs with a larger proportion of cognates are
predicted to show larger vocabulary sizes. A missing piece in this
account is how increase parallel activation might facilitate the
acquisition new words.

In this study, we hypothesise that cognates are acquired earlier in age
than non-cognates, which might explain why children learning languages
that share many cognates show larger vocabulary sizes. Evidence
supporting this claim is scarce. Bosch \& Ramon-Casas (2014) used
vocabulary parental reports (152 lexical items) from 48 Catalan-Spanish
bilinguals aged 18 months, and found that cognates represented a larger
proportion of participant's vocabulary than non-cognates. Schelletter
(2002) reported a longitudinal single case of one English-German
bilingual who produced cognates earlier than non-cognates, on average.
The low sample size in these two studies makes it challenging to draw
strong conclusions about the effect of cognateness on vocabulary growth.
On the other hand, Floccia et al.'s estimates are statically more
reliable given their (much larger) sample size, but their study was not
aimed at testing the effect of cognateness on age of acquisition
directly. In their discussion the authors state the following (pp.~70):

\begin{quote}
``This finding also provides support to the proposal that the cognate
advantage is due to cognates being acquired before non-cognates in early
childhood (Costa et al., 2016), leading to an ease of processing later
in life.''
\end{quote}

>>>>

\documentclass[
  man,
  floatsintext,
  colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue,biblatex]{apa7}

% TODO: Add custom LaTeX header directives here
\makeatletter
\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
	{0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
	{-.5em}%
	{\normalfont\normalsize\bfseries\typesectitle}}

\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{0.5em}%
	{0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
	{-\z@\relax}%
	{\normalfont\normalsize\bfseries\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
\makeatother
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}\usepackage{lineno}\linenumbers\usepackage{tipa}\usepackage[pinyin]{babel}\usepackage{CJKutf8}\usepackage{fontspec}\babelprovide[script=CJK, language={Chinese Simplified}]{chinese-simplified}\babelfont[chinese-simplified]{rm}[Ligatures = Common]{Noto Serif CJK SC}\babelfont[chinese-simplified]{sf}[Ligatures = Common]{Noto Sans CJK SC}\makeatletter\makeatother\makeatletter\makeatother\makeatletter\@ifpackageloaded{caption}{}{\usepackage{caption}}\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}\makeatother\makeatletter\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}\makeatother\makeatletter\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}\makeatother\makeatletter\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}\makeatother\makeatletter\makeatother\makeatletter\makeatother

\usepackage[style=apa,backend=biber]{biblatex}
\addbibresource{assets/references.bib}

\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Lexical frequency mediates the facilitation effect of cognateness
in bilingual lexical acquisition}
\shorttitle{Frequency and cognateness in lexical acquisition}

\authorsnames[{1},{1,2},{3},{1}]{
Gonzalo Garcia-Castro,Daniela S. Avila-Varela,Ignacio Castillejo,Núria
Sebastian-Galles
}

\authorsaffiliations{
{Center for Brain and Cognition, Universitat Pompeu Fabra},{Facultade de
Letras, Universidade de Lisboa},{Departamento de Psicología, Universidad
Autónoma de Madrid}}

\date{}
\abstract{Language non-selectivity is prominent feature of bilingual
lexical processing. An instance of such non-selectivity is embodied by
cognateness, which recent studies have suggested to facilitate
vocabulary acquisition in bilingual toddlers, who show larger vocabulary
sizes when their languages share many cognates, and to be acquired at
earlier ages than non-cognates. The specific mechanisms behind such
facilitation are unclear. We present an account of early bilingual
vocabulary growth in which cognateness interacts with lexical frequency
and language exposure to facilitate the word acquisition. We evaluated
this model against vocabulary data from 436 Catalan-Spanish bilinguals
aged 12 to 34 months. We used Bayesian Exploratory Item Response Models
to estimate participants' probability of acquisition of 604 words,
conditional to the cognate status and lexical frequency of the
word-form, and the age and degree of exposure to each language of the
toddler. We found converging evidence for an earlier age-of-acquisition
for cognate words, and for such effect being mediated by lexical
frequency and language exposure. Low-frequency words, and words from the
language of least exposure were more strongly benefitted by their
cognate status than high-frequency words. Overall, our findings support
an interactive account of bilingual vocabulary acquisition in which the
lexical representations in one language interact with the acquisition of
words in the other language.}
\addbibresource{assets/references.bib}

\keywords{lexical acquisition, vocabulary, bilingualism, item response
theory, bayesian}

\authornote{\par{\addORCIDlink{Gonzalo
Garcia-Castro}{0000-0002-8553-4209}}\par{\addORCIDlink{Daniela S.
Avila-Varela}{0000-0002-3518-8117}}\par{\addORCIDlink{Ignacio
Castillejo}{0000-0001-7445-0416}}\par{\addORCIDlink{Núria
Sebastian-Galles}{0000-0001-6938-2498}}
\par{ }
\par{   The authors declare no conflicts of interest with regard to the
funding source of this study. This research was supported by grants from
the Spanish Ministerio de Ciencia, Innovación y Universidades
(PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government
{[}ICREA (Catalan Institution for Research and Advanced Studies)
Academia 2019 award{]}. Gonzalo Garcia-Castro was supported by a
fellowship of the Spanish Ministerio de Ciencia, Innovación y
Universidades (FPI 2019). We are grateful to Chiara Santolin, Alicia
Franco-Martínez, Cristina Rodríguez-Prada, Ege E. Özer, and the rest of
the Speech Acquisition and Perception research group for their helpful
feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado
for their technical support. We also thank Cristina Dominguez and Katia
Pistrin for their efforts in recruiting infants. We also thank all
families and infants who participated in the experiments. }
\par{Correspondence concerning this article should be addressed to Gonzalo
Garcia-Castro, Center for Brain and Cognition, Universitat Pompeu
Fabra, Ramon Trias Fargas
25-27, Barcelona, Spain 8005, Email: gonzalo.garciadecastro@upf.edu}}

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, enhanced]}{\end{tcolorbox}}\fi
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The foundations of word learning are in place early in age. From six
months of age, infants start directing their gaze to some objects when
hearing their labels, according to both experimental data (Bergelson \&
Swingley, 2012, 2015; Jusczyk \& Aslin, 1995; Tincoff \& Jusczyk, 1999)
and parental vocabulary reports (Fenson et al., 2007; Samuelson, 2021).
During the last half of their second year, new word forms are acquired
at an increasingly fast rate (Bergelson, 2020; Bloom, 2002; Fenson et
al., 1994; Goldfield \& Reznick, 1990; Mayor \& Plunkett, 2011). Such
early stages of lexical acquisition are characterised by substantial
variation across children reflected, for instance, on the variability of
the number of words they know (i.e., vocabulary size, Fenson et al.,
1994; Frank et al., 2021). Despite this variability, trajectories of
early vocabulary growth seem quite stable across languages. Tardif et
al. (2008) collected data about the first ten words acquired by 10-- to
16--month-old infants living in the United States, Hong Kong, and
Beijing. Since birth, these infants had been learning English,
Cantonese, and Mandarin, respectively. The authors found a common
pattern across the three groups: their first ten words referred to
roughly the same concepts, namely relatives/caregivers (\emph{daddy},
\emph{mommy}), social routines (\emph{bye}, \emph{uh-oh}) or animal
sounds (\emph{woof-woof}). These results were later extended by Frank et
al. (2021) to a more diverse set of languages, and found that such
cross-language commonalities are stronger at earlier stages of lexical
acquisition.

Most research on early word acquisition relies extensively on data from
monolingual children, and is oblivious to substantial proportion of the
world population acquiring more than one language since early ages
(Grosjean, 2021). Previous research on bilingual vocabulary acquisition
pointed to bilingual toddlers knowing, on average, less words in each of
their languages than their monolinguals peers, and also to both groups
knowing a similar amount of words when the bilinguals' two languages are
taken into account. Hoff et al. (2012) found that English-Spanish
bilingual toddlers in South Florida (United States) knew less words in
English than monolinguals did. Both groups knew a similar total amount
of words when both English and Spanish vocabularies were counted
together, which substantiates the importance of collecting data on both
languages when assessing bilinguals' communicative development. Other
studies have provided converging evidence that bilinguals know a similar
or even larger number of words than monolinguals when the two languages
are aggregated (Gonzalez-Barrero et al., 2020; Oller \& Eilers, 2002;
Patterson, 2004; Patterson \& Pearson, 2004; Pearson et al., 1993;
Pearson \& Fernández, 1994; Petitto et al., 2001; Smithson et al.,
2014). These studies rely on bilingual samples whose languages are
relatively distant in typology: English is a Germanic language, while
Spanish is a Romance language, and both languages differ substantially
at the grammatical or the phonological level. Little is known about
vocabulary acquisition in bilinguals learning a pair of typologically
closer languages (e.g., when both are Romance languages).

For a given set of concepts, bilingual children may be exposed to two
distinct sets of word-forms: one in each language. Depending on the
similarity between both languages, the two sets of words may overlap in
form in varying degrees. When both languages are typologically close,
like Spanish and Catalan (both Romance languages), they are more likely
to share a large amount of cognates (i.e., form-similar translation
equivalents) than two typologically more distant languages, like Spanish
and English (Romance and Germanic languages, respectively). For
instance, in the presence of a door, a Spanish-Italian (or a
Spanish-Catalan) bilingual might be exposed to the words \emph{puerta}
and \emph{porta} (cognates), whereas a Spanish-English bilingual might
hear \emph{puerta} and \emph{door} (non-cognates). It could be the case
that mapping two phonologically similar labels (cognates like
\emph{puerta}-\emph{porta}) onto the same referent is easier than doing
the same with two phonologically dissimilar labels (non-cognates, like
\emph{puerta} and \emph{door}). If cognates are easier to acquire than
non-cognates, bilinguals learning two languages that share a high
proportion of cognates should benefit more often from this facilitation
effect than bilinguals learning two languages with a lower proportion of
cognates, and should therefore show larger vocabulary sizes.

Floccia, Sambrook, Delle Luche, et al. (2018) provided evidence in line
with this claim. The authors collected vocabulary data on word
comprehension and production from 372 24-month-old bilingual toddlers
living in the United Kingdom who were learning English and an additional
language. The additional language was one of 13 typologically diverse
languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek,
Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and
Welsh. The authors calculated the average phonological overlap between
the words in each of these additional languages and their translation
equivalents in English, which was taken as a proxy of the degree of
\emph{cognateness} between each pair of languages. Phonological overlap
was measured by computing the Levenshtein distance between each
cross-language pair of phonetic transcriptions. The Levenshtein distance
is a metric that computes the distance between two strings of
characters/phonemes by counting the smallest number of insertions,
deletions and substitutions one of the strings has to undergo to become
identical to the other (Levenshtein, 1966). For instance, the
Levenshtein distance for the translation equivalent
\emph{cat}-\emph{kat} (/kæt/-/\textipa{kAt}/) in English and Dutch would
be one: a single edition would be needed to make both phonological
translation identical (replacing /æ/ by /\textipa{A}/). The same
translation equivalent in English and Greek
(\emph{cat}-\emph{\(\gamma\acute{\alpha}\tau\alpha\)},
/kæt/-/\textipa{Gata}/) would score three, as it requires at least three
editions: substituting /k/ for /\textipa{G}/, substituting /æ/ for /a/
and adding another /a/.

Levenshtein distance scores are divided by the length of the longest
string to normalise the similarity scores between 0 and 1. Finally, the
result is subtracted from one, to get the complementary, and therefore a
measure of phonological similarity, instad of dissimilarity. Following
the previous examples: \(1/4 = 0.25\) for /kæt/-/\textipa{kAt}/, and
\(3/4 = 0.75\) for /kæt/-/\textipa{Gata}/. Finally, this variable was
included as a predictor of participants' vocabulary sizes. Among other
findings, Floccia and co-workers reported an increase in vocabulary size
in the additional language (i.e., not English) associated with an
increase in the average phonological similarity between the translation
equivalents of each language pair. For example, English-Dutch bilinguals
(0.2214 mean phonological overlap), were able to produce more Dutch
words than English-Mandarin bilinguals (0.0197 mean phonological
overlap) were able to produce in Mandarin.

Floccia et al.~pointed to \emph{parallel activation} as the underpinning
mechanism behind their results. The parallel activation hypothesis stems
from the language non-selective account of lexical access, which
suggests that bilinguals activate both languages simultaneously during
speech production (Costa et al., 2000; Hoshino \& Kroll, 2008) or
comprehension (Spivey \& Marian, 1999; Thierry \& Wu, 2007), even in
fully monolingual environments. One of the clearest pieces of evidence
of parallel activation was provided by Costa et al. (2000). In this
study, Spanish monolinguals and Catalan-Spanish bilingual adults were
asked to name pictures of common objects in Spanish. In half of the
trials, the object labels were cognates in Spanish and Catalan
(\emph{árbol}-\emph{arbre}, translations of \emph{tree}), whereas in the
other half of the trial labels were non-cognates
(\emph{mesa}-\emph{taula}, translations of \emph{table}), importantly
such distinction was only relevant for bilinguals. Bilinguals named
cognate pictures faster than non-cognate pictures, even after adjusting
for the lexical frequency of the items. In contrast, Spanish
monolinguals, who were unfamiliar with the Catalan translations of the
Spanish words they uttered, showed equivalent naming times for the two
types of stimuli. These results showed that bilinguals' Catalan
phonology was activated during the production of Spanish words,
facilitating the naming of cognate pictures.

Regarding word acquisition, the parallel activation account is a
plausible explanation for Floccia et al.'s results in the light of two
observations. First, bilingual toddlers' lexicon seems to consist of a
large proportion of translation equivalents: the presence of a word in
their lexicon is a good predictor of its translation equivalent also
being present (Bilson et al., 2015; Pearson et al., 1995; Tsui et al.,
2022a). Second, there is experimental evidence of parallel activation
during word comprehension in children and toddlers (Bosma \& Nota, 2020;
Poulin-Dubois et al., 2013; Von Holzen \& Mani, 2012). It is possible
that cognateness increases the amount of cross-language activation,
facilitating word acquisition. This could ultimately lead to children
who are learning languages with a larger proportion of cognates (e.g.,
English and Dutch) showing larger vocabulary sizes. This account in
which cognateness underlies the facilitation effect of typological
language distance is also in line with previous studies suggesting that
the acquisition of new words is facilitated by their phonological or
semantic similarity with other words (Fourtassi et al., 2020; Hills et
al., 2009; Jones \& Brandt, 2019; Laing, 2022). If phonological
similarity plays a cross-language facilitation role during word
acquisition, cognate translation equivalents---which share both semantic
and phonological similarity---should be acquired, on average, earlier
that non-cognate translation equivalents---which share semantic
similarity (e.g., are taxonomically related) but not phonological
similarity.

Evidence supporting an earlier age of acquisition for cognates is
scarce. Bosch and Ramon-Casas (2014) collected parental reports of
expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months,
and found that cognates represented a larger proportion of participant's
vocabulary than non-cognates. Schelletter (2002) provided converging
evidence from a longitudinal single-case study, in which an
English-German bilingual produced cognates earlier than non-cognates, on
average. The low sample size in these two studies makes it challenging
to draw strong conclusions about the effect of cognateness on vocabulary
growth. Floccia et al.'s estimates are statistically more reliable given
their remarkably larger sample size (372 participants). Although their
results suggest a cognateness effect on vocabulary size, their study was
not aimed at testing the effect of cognateness on age of acquisition at
the \emph{word-level}. The response variable in this study was the
proportion of words each participant understood and/or produced (i.e.,
vocabulary size), from the list of lexical items in the vocabulary
checklists. By aggregating the responses from all words into a single
datum per child, it was no longer possible to test the effect of
cognateness on word acquisition at the word-form level. Also, all
participants were aged 24 months, meaning that even if the unaggregated
responses to individual items were included as the response variable,
age-sensitive claims about the effect of cognateness on word acquisition
should be taken with caution (a limitation also present in the Bosch \&
Ramon-Casas, 2014 study).
